---
title: "EDA_Project"
author: "Faran Ahmad"
date: "4/11/2022"
output: html_document
---
Libraries to load before exploring data
```{r}
library(tidyverse)
library(tidyr)
library(Hmisc)
library(janitor)
library(lubridate)
library(readxl)
```

Loading Dataset and cleaning names to snake_case
```{r}
super_store_data <- readxl::read_excel("Sample - EU Superstore.xls")
super_store_data |> 
  janitor::clean_names() -> super_store_data
```

Summary_statistics
```{r}
summary(super_store_data$sales)
summary(super_store_data$profit)
```

Relation between Sales and Profit
```{r}
plotly::ggplotly(
  ggplot(super_store_data, aes(x= sales, y= profit))+
    geom_hex()+
    geom_smooth()+
    labs(x="Sales", y="Profit", title = "Relation Between Sales and Profit")
)
```

Realtion between Profit and Discount
```{r}
plotly::ggplotly(
  ggplot(super_store_data, aes(x=profit, y= discount))+
    geom_hex()+
    geom_smooth()+
    labs(x="Profit", y="Dicount", title = "Relation Between Profit and Discount")
)
```

Customers count across Segments
```{r}
plotly::ggplotly(
  ggplot(super_store_data, aes(x= segment))+
    geom_bar(fill= "steelblue")+
    labs(x= "Segments", title = "Customers count across Segments",
         y= "Frequency")
)
```

Shipment-Mode Preference
```{r}
plotly::ggplotly(
  ggplot(super_store_data, aes(x= ship_mode))+
    geom_bar(fill= "palegreen", col= "black")+
    labs(x= "Shipment mode", title = "Customers Shipment Mode preference",
         y= "Frequency")
)
```

Distribution of customers across country
```{r}
plotly::ggplotly(
  ggplot(super_store_data |> 
  group_by(country) |> count(),
  aes(x= country, y=n, fill= country))+
    geom_bar(stat = "identity")+
    labs(x="Country", fill="Labels", y="Number of Customers",
         title = "Distribution of Customers across Country")+
    theme(axis.text.x = element_text(angle = 90))
)
```

Hot selling Sub-Categories in Categories
```{r}
options(scipen = 999)
plotly::ggplotly(
  ggplot(super_store_data |> 
           group_by(category, sub_category) |> 
           summarise(sales=sum(sales)),
         aes(x=category, y= sales, fill= sub_category))+
    geom_bar(stat = "identity", col= "black")+
    labs(x= "Category", y= "Sales Distribution", fill="Sub-Category",
         title = "Hot selling Sub-Categories in Categories")
)
```

Popular Sub-Categories in Categories
```{r}
busin_orient_cat_subcat <- super_store_data |> 
  group_by(category, sub_category) |> 
  count()
options(scipen = 999)
plotly::ggplotly(
  ggplot(busin_orient_cat_subcat, aes(x= category, y= n, fill= sub_category))+
    geom_bar(stat = "identity", col= "lightyellow")+
    labs(x= "Category", y="Frequency", fill="Sub-Category",
         title = "Popular Sub-Categories in Categrory")
)
```

Sales and Profit by Sub-Category
```{r}
sales_profit_by_sub_category <- super_store_data |>
  group_by(sub_category) |> 
  summarise(sales=sum(sales),
            profit= sum(profit)) |> 
  pivot_longer(sales:profit,
               names_to = "key",
               values_to = "value")
options(scipen = 999)
plotly::ggplotly(
  ggplot(sales_profit_by_sub_category, aes(x= sub_category, y= value,
                                           fill= key))+
    geom_bar(stat = "identity", position = "dodge")+
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= "Sub-Category", y= "Total Sales and Profits",
         title = "Sales and Profit by Sub-Category")
)
```

Gross-Profit by Sub-Category
```{r}
plotly::ggplotly(
  ggplot(super_store_data |> 
           group_by(sub_category) |> 
           summarise(gross_profit= sum(profit)),
         aes(x= sub_category, y= gross_profit))+
    geom_col(fill= "lightyellow", col= "black")+
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= "Sub-Category", y= "Gross-Profit",
         title = "Gross-Profit by Sub-Category")
)
```

Gross-Profit by Category
```{r}
plotly::ggplotly(
  ggplot(super_store_data |> 
           group_by(category) |> 
           summarise(gross_profit= sum(profit)),
         aes(x= category, y= gross_profit))+
    geom_col(fill= "lightyellow", col= "black")+
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= "Category", y= "Gross-Profit",
         title = "Gross-Profit by Category")
)
```

Profit by category and subcategory
```{r}
options(scipen = 999)
plotly::ggplotly(
  ggplot(super_store_data |> 
           group_by(category,sub_category) |> 
           summarise(profit= sum(profit)),
         aes(x= category, y= profit, fill= sub_category))+
    geom_col(col="black") +
    labs(x= "Category", y= "Profit", fill= "Sub-Category",
         title = "Profit by Category and Sub-Category")
)
```

Sales Trend Over Years
```{r}
super_store_data |> 
  mutate(year= lubridate::year(order_date)) |> 
  group_by(year) |> 
  summarise(revenue= sum(sales))-> yearly_sales
options(scipen = 999)
plotly::ggplotly(
  ggplot(yearly_sales, aes(x=year, y= revenue))+
    geom_point(col="yellow", size = 2)+
    geom_line(col= "blue4") +
    labs(x= "Years", y= "Revenue", title =  "Sales Trend Over Years")
)
```

Monthly Sales Trend over Years
```{r}
super_store_data |> 
  mutate(month= lubridate::month(order_date, label = TRUE),
         year= lubridate:: year(order_date)) |>
  group_by(month, year) |> 
  summarise(sales= sum(sales)) -> monthly_sales_by_year
plotly::ggplotly(
  ggplot(monthly_sales_by_year, aes(x= month, y= sales, color= as.factor(year),
                                    group= as.factor(year)))+
    geom_line()+
    geom_point(col= "black", size= 1) +
    labs(x= "Months", y= "Sales Value", color= "Years",
         title = "Monthly Sales Trend over Years")
)
```

Monthly Purchase Frequency across years
```{r}
super_store_data |> 
  mutate(month= lubridate::month(order_date, label = TRUE),
         year= lubridate::year(order_date)) |> 
  group_by(month, year) |> 
  summarise(n_transactions= n_distinct(order_id)) ->monthly_p_freq_ov_years
plotly::ggplotly(
  ggplot(monthly_p_freq_ov_years, aes(x= month, y= n_transactions,
                                      color= as.factor(year),
                                      group= as.factor(year)))+
    geom_line()+
    geom_point(col= "black", size=1)+
    labs(x= "Months", y="Number of Transactions", color= "Years",
         title = "Montly Purchase Frequency Across Years")
)
```

Monthly Sales Trend Across Years Region Wise
```{r}
super_store_data |> 
  mutate(month= lubridate::month(order_date, label = TRUE),
         year= lubridate::year(order_date)) |> 
  group_by(month,year,region) |> 
  summarise(revenue= sum(sales)) ->m_y_sales_trend_region
plotly::ggplotly(
  ggplot(m_y_sales_trend_region, aes(x= month, y= revenue,
                                      color= as.factor(year),
                                      group= as.factor(year)))+
    geom_line()+
    geom_point(col= "black", size=1)+
    facet_wrap(~region, nrow = 3)+
    theme(axis.text.x = element_text(angle = 90))+
    labs(x= "Years", y="Sales", color= "Year",
         title = "Montly Sales Trend Across Years Region Wise")
)
```

Average Life-Span of Customers with Business in Months
```{r}
super_store_data |>
  group_by(customer_id) |> 
  summarise(monthly_lifespan= difftime(max(order_date), min(order_date), units = "days")) |>
  mutate(monthly_lifespan= round(as.integer(monthly_lifespan/30))) -> cus_eng_monthly
plotly::ggplotly(
  ggplot(cus_eng_monthly |> 
           filter(monthly_lifespan > 0),
         aes(x= monthly_lifespan))+
    geom_bar(fill= "steelblue", col= "lightyellow")+
    labs(x= "Monthly Life-Span", y= "Number of Customers",
         title = "Average Life-Span of Customers with Business in Months")
)
```

Dominant Sub-Categories Over Years (Transactions)
```{r}
super_store_data |> 
  mutate(year= lubridate::year(order_date)) |> 
  group_by(sub_category, year) |> 
  summarise(n_transactions= n_distinct(order_id)) -> dom_sbcat_o_years
plotly::ggplotly(
  ggplot(dom_sbcat_o_years, aes(x= sub_category, y= n_transactions,
                                color= as.factor(year),
                                group= as.factor(year)))+
    geom_line()+
    labs(x= "Number of Transactions", y= "Sub-Category",
         title = "Dominant Sub-Categories Over Years (Transactions)") +
    theme(axis.text.x = element_text(angle = 45))
)
```

Dominant Category Over Years (Transactions)
```{r}
super_store_data |> 
  mutate(year= lubridate::year(order_date)) |> 
  group_by(category, year) |> 
  summarise(n_transactions= n_distinct(order_id)) -> dom_cat_o_years
plotly::ggplotly(
  ggplot(dom_cat_o_years, aes(x= category, y= n_transactions,
                              color= as.factor(year), group= as.factor(year))) +
    geom_line() +
    geom_point(col= "black", size = 1)+
    labs(x= "Category", y= "Number of Transactions", color= "Year",
         title = "Dominant Sub-Categories Over Years (Transactions)")
)
```

Dominant Segment Over Years (Revenue)
```{r}
super_store_data |> 
  mutate(year= lubridate::year(order_date)) |> 
  group_by(segment, year) |> 
  summarise(revenue= sum(sales)) -> dom_seg_o_years
plotly::ggplotly(
  ggplot(dom_seg_o_years, aes(x= segment, y= revenue, color= as.factor(year),
                              group= as.factor(year))) +
    geom_line()+
    geom_point(col= "black", size= 1) +
    labs(x= "Segment", y= "Revenue",
         title = "Dominant Segment Over Years (Revenue)")
)
```

Dominant Country Over Years (Profit)
```{r}
super_store_data |> 
  mutate(year= lubridate::year(order_date)) |> 
  group_by(country, year) |> 
  summarise(profit= sum(profit)) -> dom_country_o_years
plotly::ggplotly(
  ggplot(dom_country_o_years, aes(x= country, y= profit, col= as.factor(year),
         group= as.factor(year)))+
    geom_line()+
    geom_point(col= "black", size= 1)+
    labs(x= "Country", y= "Profit",
         title = "Dominant Country Over Years (Profit)")+
    theme(axis.text.x = element_text(angle = 90))
)
```

Average Basket size of Sub-Categories
```{r}
super_store_data |> 
  mutate(n_transactions= n_distinct(order_id)) |> 
  group_by(sub_category) |> 
  summarise(quantity= sum(quantity),
            n_transactions= n_transactions) |> 
  distinct() |> 
  mutate(basket_size= quantity/n_transactions) -> avg_basket_size_subcat
plotly::ggplotly(
  ggplot(avg_basket_size_subcat, aes(x= basket_size, y= sub_category))+
    geom_col(fill="darkseagreen", col="black")+
    labs(x= "Average Basket Size", y= "Sub-Category",
         title = "Average Basket size of Sub-Categories")
)
```

Average Basket size of Sub-Categories By Region
```{r}
super_store_data |> 
  mutate(n_transactions= n_distinct(order_id)) |> 
  group_by(sub_category,region) |> 
  summarise(quantity= sum(quantity),
            n_transactions= n_transactions) |>
  mutate(basket_size= quantity/n_transactions) ->avg_basket_size_region
plotly::ggplotly(
  ggplot(avg_basket_size_region, aes(x= basket_size, y= sub_category))+
    geom_col(fill="darkseagreen") +
    facet_wrap(~region) +
    labs(x= "Average Basket Size", y= "Sub-Category",
         title = "Average Basket size of Sub-Categories By Region")
)
```

Average Basket size by product
```{r}
super_store_data |> 
  mutate(n_transactions= n_distinct(order_id)) |> 
  filter(sub_category == "Tables") |> 
  group_by(product_id,sub_category) |> 
  summarise(quantity= sum(quantity),
            n_transactions= n_transactions) |>
  distinct() |>
  mutate(basket_size= quantity/n_transactions) ->avg_basket_size_product
plotly::ggplotly(
  ggplot(avg_basket_size_product, aes(x= basket_size, y= product_id)) +
    geom_col(fill= "steelblue", col= "black") +
    facet_wrap(~ sub_category) +
    labs(x= "Average Basket-Size", y= "Product-ID",
         title = "Average Basket-Size by Product")
)
```

Purchase Frequency Across Week-Days in Months
```{r}
super_store_data |> 
  mutate(month= lubridate::month(order_date, label = TRUE),
         week_day=weekdays(order_date)) |> 
  group_by(month, week_day) |>
  summarise(n_transactions= n_distinct(order_id)) ->n_trans_week_days
plotly::ggplotly(
  ggplot(n_trans_week_days, aes(x= month, y= week_day, fill= n_transactions))+
    geom_tile()+
    labs(x= "Month", y= "Week-Days", fill = "Purchase Frequency",
         title = "Purchase Frequency Across Week-Days in Months")
)
```

RFM-Analysis
```{r}
date_today <- lubridate::ymd("2019/04/10")
super_store_data |> 
  group_by(customer_id) |> 
  summarise(total_revenue= sum(sales),
            n_transactions= n_distinct(order_id),
            last_purchase_date= max(order_date)) |>
  mutate(days_since_last_purchase=
           difftime(date_today, last_purchase_date, units = "days")) |>
  filter(total_revenue > 0) |> 
  mutate(monetary_quintile = cut2(total_revenue, g =5),
         frequency_quintile = cut2(n_transactions, g = 5),
         recency_quintile = cut2(days_since_last_purchase, g = 5)) |> 
  mutate(recency_score = as.integer(recency_quintile),
         frequency_score = as.integer(frequency_quintile),
         monetary_score = as.integer(monetary_quintile)) |> 
  mutate(recency_score = dense_rank(desc(recency_score))) |>
  mutate(RFM= recency_score * 100 + frequency_score * 10 + monetary_score) |>
  mutate(segments= ifelse(recency_score >= 4 & frequency_score >= 3 & monetary_score >= 4, "Champions",
                          ifelse((recency_score >= 4) & (frequency_score <= 2) & (monetary_score >= 4), "High Spending New Customers",
                                 ifelse((recency_score >= 4) & (frequency_score >= 4) & (monetary_score == 3), "Average Spending Champions",
                                        ifelse((recency_score >= 2 & recency_score <= 4) & (frequency_score >= 3 & frequency_score <= 5) & (monetary_score >= 4), "Loyal Customers", 
                                               ifelse((recency_score >= 3) & (frequency_score >= 1 & frequency_score <= 3) & (monetary_score >= 1 & monetary_score <= 3), "Potential Loyalists",
                                                      ifelse((recency_score >= 4 & recency_score <= 5) & (frequency_score < 2) & (monetary_score < 2), "New Customers",
                                                             ifelse((recency_score >= 3 & recency_score <= 4) & (frequency_score < 2) & (monetary_score < 2), "Promising",
                                                                    ifelse((recency_score >= 3 & recency_score <= 4) & (frequency_score >= 2 & frequency_score <= 4) & (monetary_score >= 3 & monetary_score <= 5), "Need attention",
                                                                           ifelse((recency_score >= 2 & recency_score <= 3) & (frequency_score < 3) & (monetary_score < 3), "About to sleep",
                                                                                  ifelse((recency_score < 3) & (frequency_score >=2 & frequency_score <= 5) & (monetary_score >= 2 & monetary_score <= 5), "At risk",
                                                                                         ifelse((recency_score < 2) & (frequency_score >= 4 & frequency_score <= 5) & (monetary_score >= 4 & monetary_score <= 5), "Can't loose them",
                                                                                                ifelse((recency_score >= 2 & recency_score <=3) & (frequency_score >= 2 & frequency_score <= 3) & (monetary_score >= 2 & monetary_score <= 3), "Hibernating",
                                                                                                       ifelse((recency_score <= 2) & (frequency_score <= 2) & (monetary_score >= 4), "High Value Lost",
                                                                                                              ifelse((recency_score < 2) & (frequency_score <= 3) & (monetary_score <= 2), "Low Value Lost",
                                                                                                                     ifelse((recency_score == 3) & (frequency_score < 2) & (monetary_score >= 4), "High Spending New Customers",
                                                                                                                            ifelse((recency_score <= 2) & (frequency_score < 2) & (monetary_score == 3), "Average Spending Lost",
                                                                                                                                   ifelse((recency_score <= 2) &(frequency_score <= 4) &(monetary_score == 1), "Low Value Hibernating",
                                                                                                                                          ifelse((recency_score <= 3) &(frequency_score >= 4) &(monetary_score <=3), "Average Spending Need Attention", "Low Spending Champions"))))))))))))))))))) -> rfm
```

Number of Customers in Segments
```{r}
plotly::ggplotly(
  ggplot(rfm, aes(x= segments, fill= segments))+
    geom_bar(col= "black")+
    theme(axis.text.x = element_text(angle = 45))
)
```

Calculating CLTV
```{r}
# purchase value
super_store_data |> 
  group_by(order_id) |> 
  summarise(purchase_value= sum(sales)) |> 
  ungroup() |> 
  summarise(avg_purch_value= mean(purchase_value)) -> avg_purchase_value

# purchase frequency        
super_store_data |> 
  mutate(month= lubridate::month(order_date),
         year= lubridate::year(order_date)) |> 
  group_by(customer_id, month,year) |> 
  summarise(purchase_frequency= n_distinct(order_id)) |>
  ungroup() |> 
  summarise(avg_purch_frequency= mean(purchase_frequency)) ->avg_purchase_frequency

#Gross margin
super_store_data |> 
  mutate(gross_margin= sum(profit)/ sum(sales))
gross_margin <- 0.1268953

# life span
super_store_data |>
  mutate(month= lubridate::month(order_date)) |> 
  group_by(customer_id) |> 
  summarise(lifespan= difftime(max(order_date), min(order_date), units = "days")) |> 
  mutate(lifespan= round(as.integer(lifespan/30))) |>
  summarise(avg_life_span= mean(lifespan)) -> avg_life_span


CLTV= (avg_purchase_value$avg_purch_value) * (gross_margin) * (avg_purchase_frequency$avg_purch_frequency) * (avg_life_span$avg_life_span)
CLTV
```

JOINS
Ship-Mode Preference by Segments (JOINS)
```{r}
left_join(rfm, super_store_data) -> joined
plotly::ggplotly(
  ggplot(joined |> 
           group_by(segments) |> 
           count(ship_mode), aes(x= segments, y= n,
                                 fill= ship_mode))+
    geom_col(position = "fill", col= "black")+
    theme(axis.text.x = element_text(angle = 90))+
    labs(x= "Customer Segments", y= "Count", fill= "Shipment-Mode",
         title = "Ship-Mode Preference by Segments")+
    coord_flip()
)
```

Segments Count Across Sub-Categories
```{r}
plotly::ggplotly(
  ggplot(joined |> 
           group_by(sub_category) |> 
           count(segments), aes(x= segments, y= n, fill= sub_category))+
    geom_col(position = "fill", col="black") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x= "Customer Segments", y= "Count", fill= "Sub-Category",
         title = "Segments Count Across Sub-Categories")
)
```

